\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-16 tex/paper.tex
save

We consider a family of parametrized densities $p(\mathbf{x}|\bftheta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\bftheta$ are the
parameters of interest. The simulation may involve some complicated latent
process
%, such that
%\begin{equation}\label{eqn:p_x}
%    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
%\end{equation}
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
where $\bfz \in {\cal Z}$ is a latent variable providing an external
source of randomness.
Unlike implicit generative models defined by neural networks, we do not assume
$\bfz$ to be a fixed-size vector with a simple density. Instead, the
dimension of $\bfz$ and the nature of its components (uniform, normal,
discrete, continuous, etc.) are inherited from the control flow of the
simulation code and may depend on $\bftheta$ in some intricate way. Moreover,
the dimension of $\bfz$ may be much larger than the dimension of
$\bfx$. Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.

@@

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
parameters of $g$ and $p(\mathbf{x}|\bftheta)$ and alternately minimize the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\bfphi$ and $\bfpsi$.
When updating
$\bfphi$, unbiased estimates of $\nabla_\bfphi U_d$ can be obtained by
directly evaluating the gradient of $U_d$ over mini-batches of real and
generated data, as is ordinarily done in stochastic gradient descent. When updating
$\bfpsi$, $\nabla_\bfpsi U_g$ can be estimated as described in the previous section.
That is,
%\begin{equation}\label{eqn:grad-ug-approx}
%   \nabla_\bfpsi U_g = \mathbb{E}_{\subarray{l}{\bftheta \sim q(\bftheta|\bfpsi), \\ \hat{\bfx} \sim p(\bfx | \bftheta) }}  [-d( \hat{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
%\end{equation}
\begin{equation}\label{eqn:grad-ug-approx}
   \nabla_\bfpsi U_g = \mathbb{E}_{\substack{\bftheta \sim q(\bftheta|\bfpsi), \\ \tilde{\bfx} \sim p(\bfx | \bftheta)}}  [-d( \tilde{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
\end{equation}
%\begin{equation}\label{eqn:grad-ug-approx}
%    \nabla_\bfpsi U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta)}  [-d(g(\bfz;\bftheta);\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
%\end{equation}
which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \bfphi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
\end{equation}
for $\bftheta_m \sim q(\bftheta|\bfpsi)$ and $\bfz_m \sim p(\bfz|\bftheta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization (AVO) procedure, as built on top of WGAN-GP.
%
%Obviously, the variational relaxation could similarly be coupled with
%other variants of GANs and/or of evolution strategies.

@@

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems. Improvements are possible using variance reduction techniques such as Rao-Blackwellization,
Control Variates, and quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
Moreover, the algorithm could be adopted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic. These variants may be particularly useful for small $N$.

@@

We can also view the optimization  through the lens of empirical Bayes (EB),
where one specifies a family of priors and $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
%In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the
%posterior predictive distribution.
Traditionally, we would have access to the likelihood and optimize with respect to the family of posteriors
 $p(\bftheta | \bfx, \bfpsi) = p(\bfx | \bftheta) p(\bftheta | \bfpsi) / \int p(\bfx | \bftheta) p(\bftheta | \bfpsi) d\bftheta $
as an intermediate step to computing $\bfpsi^*$. In our case, the likelihood is intractable and we have not explicitly
specified a prior family. However, the procedure corresponds to empirical Bayes if we identify
$q(\bftheta | \bfpsi^*) = p(\bftheta | \bfx , \bfpsi)$. Thus we call our method AVO-EB and note that it is applicable even outside of the likelihood-free context.

Algorithm~\ref{alg:avo}, can be easily modified to have different mini-batch $M_r$ for the observed data and $M_g$ for the simulator.
When $M_g \gg M_r$, then the variance of $\bftheta^*$ scales ad $\sqrt{M_r}$~\citep{bernton2017inference}. A final $\sqrt{N/M}$ scaling to the log-posterior $q(\bftheta | \bfpsi)$ is needed to reflect the full strength of the data.

>>>>

We consider a family of parametrized densities $p(\mathbf{x}|\bftheta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\bftheta$ are the
parameters of interest. The simulation may involve some complicated latent
process
%, such that
%\begin{equation}\label{eqn:p_x}
%    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
%\end{equation}
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
where $\bfz \in {\cal Z}$ is a latent variable providing an external
source of randomness.
Unlike implicit generative models defined by neural networks, we do not assume
$\bfz$ to be a fixed-size vector with a simple density. Instead, the
dimension of $\bfz$ and the nature of its components (uniform, normal,
discrete, continuous, etc.) are inherited from the control flow of the
simulation code and may depend on $\bftheta$ in some intricate way. Moreover,
the dimension of $\bfz$ may be much larger than the dimension of
$\bfx$.
%Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.

@@

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
parameters of $g$ and alternately minimize the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\bfphi$ and $\bfpsi$.
When updating
$\bfphi$, unbiased estimates of $\nabla_\bfphi U_d$ can be obtained by
directly evaluating the gradient of $U_d$ over mini-batches of real and
generated data, as is ordinarily done in stochastic gradient descent. When updating
$\bfpsi$, $\nabla_\bfpsi U_g$ can be estimated as described in the previous section.
That is,
%\begin{equation}\label{eqn:grad-ug-approx}
%   \nabla_\bfpsi U_g = \mathbb{E}_{\subarray{l}{\bftheta \sim q(\bftheta|\bfpsi), \\ \hat{\bfx} \sim p(\bfx | \bftheta) }}  [-d( \hat{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
%\end{equation}
\begin{equation}\label{eqn:grad-ug-approx}
   \nabla_\bfpsi U_g = \mathbb{E}_{\substack{\bftheta \sim q(\bftheta|\bfpsi), \\ \tilde{\bfx} \sim p(\bfx | \bftheta)}}  [-d( \tilde{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
\end{equation}
%\begin{equation}\label{eqn:grad-ug-approx}
%    \nabla_\bfpsi U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta)}  [-d(g(\bfz;\bftheta);\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
%\end{equation}
which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \bfphi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
\end{equation}
for $\bftheta_m \sim q(\bftheta|\bfpsi)$ and $\bfz_m \sim p(\bfz|\bftheta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization (AVO) procedure, as built on top of WGAN-GP.
%
%Obviously, the variational relaxation could similarly be coupled with
%other variants of GANs and/or of evolution strategies.

@@

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems. Improvements are possible using variance reduction techniques such as Rao-Blackwellization,
Control Variates, and quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic. These variants may be particularly useful for small $N$.

@@

We can also view the optimization  through the lens of empirical Bayes (EB),
where one specifies a family of priors $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
%In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the
%posterior predictive distribution.
Traditionally, we would have access to the likelihood and optimize with respect to the family of posteriors
 $p(\bftheta | \bfx, \bfpsi) = p(\bfx | \bftheta) p(\bftheta | \bfpsi) / \int p(\bfx | \bftheta) p(\bftheta | \bfpsi) d\bftheta $
as an intermediate step to computing $\bfpsi^*$. In our case, the likelihood is intractable and we have not explicitly
specified a prior family. However, the procedure corresponds to empirical Bayes if we identify
$q(\bftheta | \bfpsi^*) = p(\bftheta | \bfx , \bfpsi)$. Thus we call our method AVO-EB and note that it is applicable even outside of the likelihood-free context.

Algorithm~\ref{alg:avo}, can be easily modified to have different sizes of mini-batches,
$M_r$ for the observed data and $M_g$ for the simulator. When $M_g \gg M_r$,
then the variance of $\bftheta^*$ scales as
$\sqrt{M_r}$~\citep{bernton2017inference}. A final $\sqrt{N/M}$ scaling to the
log-posterior $q(\bftheta | \bfpsi)$ is needed to reflect the full strength of
the data.
